A NetdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm

### 2024-12-05 example.qmd
Working on the disadoption

This paper provides a general overview of the R package, netdiffuseR, a library for estimating contagion and diffusion processes on networks. We provide a brief introduction to diffusion of innovations theory, introduce netdiffuseR, provide a tutorial for empirical work, and examples of simulations. We also describe ways to analyze multiple simultaneous behavior diffusion such as adoption of complementary or competing products. Lastly we demonstrate how to model disadoption. This paper demonstrates how to use the netdiffuseR package to read, analyze, visualize, and simulate diffusion networks.

Diffusion of innovations is a theoretical approach that explains how new ideas and practices spread within and between communities. Diffusion of innovations theory has been used to study the spread of: fertility preferences and contraceptive choices (Bongaarts & Watkins, 1996; Valente et al., 1997); risk behaviors such as smoking (Christakis & Fowler, 2007); alcohol use (Light, et al., 2013); substance use (Valente et al., 2006; 2013); physician behavior (Coleman et al., 1966; Iyengar, et al., 2011); public health policies (Valente, et al., 2015; Yamagata et al., 2017); physical activity (Aral & Nicolaides, 2017); among other behaviors. Thus, developing a platform to easily analyze and compare factors associated with network influences on adoption has broad applicability in the behavioral and physical sciences.

Diffusion theory states that a new idea or practice enters a community or population from some external source such as an impersonal medium, a person from another community, or a technological shift (Rogers, 2003; Valente, 1995).  The idea or practice then spreads from person to person through interpersonal communication. Thus, the theory and techniques of social network analysis are integral to the study of the diffusion of innovations (Valente, 2010), and constitute the field of diffusion networks.

@@

Where $W_{ij}$ is a weight matrix indicating ties between individuals, most commonly a binary adjacency matrix such as who is friends with whom but can be any relation; a is an adoption vector indicating which of i’s j alters had adopted at time t-1; and the denomitor is the row sum, the number of outgoing ties. Eit in (1) measures the proportion of a person’s ties at time t who have adopted the behavior at the prior time period, which is a contagion measure. As described in Valente (2024), however, there can be many extensions of the exposure calculation by estimating influence of ties based on: a count (removing the denominator), indirect connectivity, structural equivalence, strength (weak vs. strong), Simmelian relations (alter-to-alter), alter network properties such as centrality, etc. NetdiffuseR enables users to calculate these alternative exposures with a parameter toggle in the exposure function.

There are other measures needed to estimate diffusion network effects. Network thresholds (Valente, 1996) are the proportion or number of adopters in one’s network when an individual adopts. Some people innovate with none or few adopters in their network adopting whereas others wait until a majority or all adopt. Network thresholds provide a means to estimate the classic two-step flow hypotheses (Katz & Lazarsfeld, 1957) and have been shown to moderate health promotion impacts (Valente & Saba, 1998). The threshold distribution drives the diffusion rate and prevalence (Valente & Vega Yon, 2020). Other diffusion network measures are Moran’s I, a measure of contagion, network susceptibility, network influence (Valente, et al, 2015), the hazard rate, and Bass model parameter coefficients (Bass, 1969; Valente, 1993).

@@

netdiffuseR also can handle adoption processes for more than one innovation/behavior. To select a multiadoption process, you only must give a `list` as the `seed.p.adopt` parameter. Therefore the simplest way to run `rdiffnet` with multiple behaviors could be:

@@

Now, the characteristics of the behaviors being simulated can be given together, all sharing the same specifications,

```{r}
#| label: simulation-multiadoption-process-5
diffnet_5 <- rdiffnet(
  200, 10,
  seed.p.adopt   = list(0.1, 0.15),
  threshold.dist = runif(200, .3,.5),
  seed.nodes     = 'central', 
  behavior       = 'tobacco'
  )

diffnet_5
```

and in that case, for two behaviors, they are named `‘behavior_name’`_1 and `‘behavior_name’`_2; or you can specify different characteristics for each behavior being simulated:

```{r}
#| label: simulation-multiadoption-process-6
diffnet_6 <- rdiffnet(
  200, 10,
  seed.p.adopt   = list(0.1, 0.15),
  threshold.dist = list(
    runif(200, .3,.5),
    runif(200, .2,.4)
  ),
  seed.nodes     = list('central', 'random'),
  behavior       = list('tobacco', 'alcohol')
 )

@@

In this particular example, we ran `rdiffnet` with most of the parameters being `list`. The reader is invited to look at other types of possible inputs running `?rdiffnet`.

Same as before, you can give a specific network as input. Besides passing fixed networks as we did with the small-world example one, the `rdiffnet` function also supports passing diffnet objects as input. When doing so, the function will use the graph of the diffnet object as the seed graph and will take the time argument as the number of timepoints included in the graph:

```{r}
#| label: simulation-multiadoption-process-7
# Run the simulation with the same parameters as before
# but specifying different initial proportion of adopters
# for each behavior
diffnet <- rdiffnet(
  seed.graph     = kfam_diffnet,
  seed.p.adopt   = list(0.1, 0.15),
  threshold.dist = runif(nvertices(kfam_diffnet), .3,.5)
  )

>>>>

This paper provides a general overview of the R package, netdiffuseR, a library for estimating contagion and diffusion processes on networks. We provide a brief introduction to diffusion of innovations theory, introduce netdiffuseR, provide a tutorial for empirical work, and examples of simulations. We also describe ways to analyze multiple simultaneous behavior diffusion such as adoption of complementary or competing products. Lastly we demonstrate how to model disadoption. This paper enable researchers to use the netdiffuseR package to read, analyze, visualize, and simulate diffusion networks.

Diffusion of innovations is a theoretical approach that explains how new ideas and practices spread within and between communities. Diffusion of innovations theory has been used to study the spread of: fertility preferences and contraceptive choices (@bongaarts_social_1996; @valente_social_1997); risk behaviors such as smoking (@christakis_spread_2007); alcohol use (@light_onset_2013); substance use (@hoffman_peer_2006; @fujimoto_alcohol_2013); physician behavior (@Coleman1966; @iyengar_opinion_2011); public health policies (@valente_diffusion_2015; @yamagata2017); physical activity (@aral_exercise_2017); among other behaviors. Thus, developing a platform to easily analyze and compare factors associated with network influences on adoption has broad applicability in the behavioral and physical sciences.

Diffusion theory states that a new idea or practice enters a community or population from some external source such as an impersonal medium, a person from another community, or a technological shift (@rogers_diffusion_2003; @valente_origins_1995).  The idea or practice then spreads from person to person through interpersonal communication. Thus, the theory and techniques of social network analysis are integral to the study of the diffusion of innovations (@valente_social_2010), and constitute the field of diffusion networks.

@@

Where $W_{ij}$ is a weight matrix indicating ties between individuals, most commonly a binary adjacency matrix such as who is friends with whom but can be any relation; a is an adoption vector indicating which of i’s j alters had adopted at time t-1; and the denominator is the row sum, the number of outgoing ties. Eit in (1) measures the proportion of a person’s ties at time t who have adopted the behavior at the prior time period, which is a contagion measure. As described in @Valente2024, however, there can be many extensions of the exposure calculation by estimating influence of ties based on: a count (removing the denominator), indirect connectivity, structural equivalence, strength (weak vs. strong), homophily, Simmelian relations (alter-to-alter), alter network properties such as centrality, etc. NetdiffuseR enables users to calculate these alternative exposures with a parameter toggle in the exposure function.

There are other measures needed to estimate diffusion network effects. Network thresholds (@valente_threshold_1996) are the proportion or number of adopters in one’s network when an individual adopts. Some people innovate with none or few adopters in their network adopting whereas others wait until a majority or all adopt. Network thresholds provide a means to estimate the classic two-step flow hypotheses (Katz & Lazarsfeld, 1957) and have been shown to moderate health promotion impacts (Valente & Saba, 1998). In addition, the threshold distribution drives the diffusion rate and prevalence (Valente & Vega Yon, 2020). Other diffusion network measures are Moran’s I, a measure of contagion, network susceptibility, network influence (Valente, et al, 2015), the hazard rate, and Bass model parameter coefficients (Bass, 1969; Valente, 1993).

@@

*netdiffuseR* also can handle adoption processes for more than one behavior. To select a multiadoption process, you can pass a `list` as the `seed.p.adopt` parameter. Here is a simple example using the Koren Family Planning network:

@@

Inspecting the output from the print method of the `diffnet` object, we can see that the object contains two behaviors: The  "Num of behaviors" entry now shows `2`, the "Behavior" entry also shows two behaviors, `"Random contagion \_1, Random contagion\_2"`, and finally, the "Prevalence" entry also shows two numbers: `0.92, 0.96`. Although we are simulating two behaviors, `rdiffnet` will simulate as many as values are in the `seed.p.adopt` list.

::: {.callout-tip}
In the current implementation of `rdiffnet`, the multiadoption module simulates behaviors independently. That is, the code above would be equivalent to simulating the same behavior twice. More complicated models in which behaviors are interdependent are supported via the disadoption parameter.
:::

`rdiffnet`'s defaults will replicate the simulation parameters across behaviors. Nonetheless, we can use lists to specify different parameters for each behavior. For example, the following code simulates two behaviors with different initial adopters, threshold distributions, seed nodes, and labels for the behaviors:

```{r}
#| label: simulation-multiadoption-process-6
diffnet_6 <- rdiffnet(
  200, 10,
  seed.p.adopt = list(0.1, 0.15),
  threshold.dist = list(
    runif(200, .3, .5),
    runif(200, .2, .4)
  ),
  seed.nodes = list("central", "random"),
  behavior   = list("tobacco", "alcohol")
 )

@@

In this particular example, we ran `rdiffnet` with most of the parameters being `list`. The reader is invited to look at other types of possible inputs in the `rdiffnet` documentation.

As we did in the previous section, we can give a specific network as input. Besides passing fixed networks as we did with the small-world example, the `rdiffnet` function also supports passing diffnet objects as input. When doing so, the function will use the graph of the diffnet object as the seed graph and will take the time argument as the number of timepoints included in the graph:

```{r}
#| label: simulation-multiadoption-process-7
diffnet <- rdiffnet(
  seed.graph     = kfam_diffnet,
  seed.p.adopt   = list(0.1, 0.15),
  threshold.dist = runif(nvertices(kfam_diffnet), .3,.5)
  )
